A generalization of the Bradley-Terry model for draws in chess

### 2019-05-27 inst/collusion_chess_draws.Rnw
minor rewording

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be two-way
competitive situations such as tennis in which one player is
identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
considers~$n$ players and assigns non-negative number $p_i$ to player
$i$, with $\sum p_i=1$.  Then player $i$ beats player $j$ with
probability $\frac{p_i}{p_i+p_j}$.  This simple system has been
modified and extended many times, for example to account for multi-way
competitions such as {\em MasterChef Australia}~\citep{hankin2017},
and non-competitive situations such as consumer
behaviour~\citep{hankin2010} and mating preference~\citep{west2008}.

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage
winning about 37.5\% of tournament games (black wins 27.6\%); and
secondly, many chess games, about 39.4\%, are drawn (figures from
Chessbase).

\subsection{Draws} Chess games may be won by either player, or may be
drawn; the implication of a draw is that neither player is better than
the other.  Before 1867, tournament games that were drawn were
replayed until one or other player won---reducing a chess game to a
Bernoulli trial (with probability one).  This system became unworkable
due to the large number of draws, resulting in the modern practice of
awarding each player of a drawn game half a point, keeping chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are usually poorer in spectacular or striking elements such as
sharp tactical combinations or sacrifice.  Excessive drawing dissuades
sponsors, leaves spectators feeling cheated, and gives commentators
less material to discuss.  Players themselves do not seem to value
draws in the wider context of the game: \citet{fischer1969}, for
example, includes only 8 draws in his compilation of 60 memorable
games, surely not representative of tournament play in which about
39.4\% of games are drawn (binomial test, $p<10^{-5}$).

@@

Such arguments suggest that one should infer more from a drawn game
than simple equality of skill: the play itself tends to be different
and more cautious, the drawing players have a different mindset, and
drawn games have less value in the wider context of the game.

\subsection{First-mover advantage} Traditionally White moves first in
any game, and this feature introduces an asymmetry into play.
Historically, we see about 1.35 white wins for every black win.
However, interpretations of this statistic vary and the issue is hotly
debated with opinions ranging from that of~\cite{adams2007}---``White
to play and win'' to~\citet{adorjan1988}---``Black is OK!''.

@@

Here I show how Bradley-Terry may be generalized to account for the
existence of draws and the first-mover advantage.  Consider first the
standard Bradley-Terry model for a finite number of chess players.
Ignoring the effect of playing white, we wish to find numbers
(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant
0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is
$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
assumption of independence of trials, would be

@@

where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess game is regarded as a
three-way competition between the two players and a third entity (a
``draw monster'') that wins if the game is drawn.  The system is
further modified to account for first-move advantage by positing an
entity whose strength is added to that of the white player.  This
methodology affords insight into players' strengths, response to
playing black or white, and risk-aversion as manifested by probability
of drawing.  If we consider a chess game to be a three-way trial, an
appropriate likelihood function would be for non-negative
$p_1,\ldots,p_n,\draw$ (here \draw\ is the draw monster's strength,
fraktur font being used for reified entities' strengths) with $\draw+\sum
p_i=1$ and

@@

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength \white\ is
added to the white player.  The strength of the white player is thus
the sum of the player's strength and \white.  Recalling that a game
might be a draw, we obtain the probabilities as given in
Table~\ref{probabilities}.  The corresponding likelihood function is
given below in equation \ref{likelihood_WD}:

@@

\begin{align}\nonumber
\prod_{i<j}&{}
\left(\frac{p_i+\white}{p_i+p_j+\draw+\white}  \right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+\draw+\white    }\right)^{L_{ij}^\mathrm{white}}\\
&{} \left(\frac{p_i}{p_i+p_j+\draw+\white}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
\left(\frac{p_j+\white}{p_i+p_j+\draw+\white  }\right)^{L_{ij}^\mathrm{black}}\\
&{} \left(\frac{\draw}{p_i+p_j+\draw    }\right)^{D_{ij}}
\nonumber
\end{align}

where $W_{ij}^\mathrm{white}$ is the number of games won by $i$ when
playing white against $j$, and so on.  The final term is algebraically
equal to

\begin{equation}
\prod_{i<j}
\left(\frac{\draw}{p_i+p_j+\draw+\white}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{\draw}{p_i+p_j+\draw+\white}\right)^{D_{ij}^\mathrm{black}}
=
\prod_{i<j}
\left(\frac{\draw}{p_i+p_j+\draw}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
\end{equation}

which might be preferable on grounds of clarity.  We now apply this
and related likelihood models to real chess datasets using the Method
of Support~\citep{edwards1972}.  To test any hypothesis $H$ on
likelihood function $\mathcal{L}$ one first identifies the constraint
on hypothesis space corresponding to $H$.  One then finds the
unconstrained maximum likelihood $\mathcal{L}_\mathrm{free}$ and
compares it with the maximum likelihood $\mathcal{L}_\mathrm{cons}$
when the optimization is constrained by $H$.  The test statistic is
then the likelihood ratio
$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
$n$ is the number of degrees of freedom implemented by $H$.

@@

The Method of Support \citep{edwards1972} involves only numerical
optimization (not integration which is computationally expensive).
The analysis presented here uses the framework of the
R~\citep{rcore2019} \pkg{hyper2} package~\citep{hankin2017} as a
convenient and unified interface.  Because gradients are available,
convergence to the evaluate is generally rapid and hypothesis-specific
constraints on parameter space are, in general, linear and
straightforward to implement.

@@

Grandmasters Anand, Karpov, and Kasparov have long-standing rivalry
and have played many games amongst themselves.  A summary may be found
in table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes

@@

\begin{figure}[t]
\begin{center}
<<ddc,echo=FALSE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
  \end{center}
\end{figure}

This likelihood function is easily maximized numerically and the
evaluate is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $\draw=0$ cannot possibly be the
case: the likelihood for this hypothesis would be identically zero,
owing to the nonzero term in the numerator of
equation~\ref{kka_likelihood}.  From the figure, we see that the
maximum likelihood estimate for draw strength $\hat{\draw}\simeq 0.53$:
compare the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 games.  The na\"{i}ve estimate is lower
because the maximization process automatically adjusts for the
different players' strengths.  Secondly, we may make inferences about
$\white$, the extra strength conferred by playing white: it is conceivable
that $\white=0$ as this term does not appear on its own in the numerator of
any term.  We may test $H_0\colon \white=0$ using the Method of Support by
maximizing the likelihood subject to the constraint $\white=0$ and
comparing with the likelihood when the optimization is free.

@@

As argued above, players with a cautious style tend to draw and this
is an important and interesting component of the game.  To investigate
this we need a likelihood function that can account for possible
differences in draw proclivity among the players.  This may be
achieved by considering a game between two players to be a contest
between five entities: the two real players, the white advantage, and
two {\em personalised} draw entities, one for each real player; a
drawn game is counted as a win for either of the draw monsters.  Under
these assumptions, table~\ref{kka} translates into the following
likelihood function:

@@

Here I consider game results from Cura\c{c}ao 1962, reproduced in
Table~\ref{candidates}.  We see that all twelve of the games between
the three accused players ended in a draw, a significant result given
51 winning games and 54 draws (Fisher's exact test, one-sided,
$p=0.0079$).  To quantify such Soviet collusion, accounting for the
player skills, we introduce a new draw entity: a collusive draw
monster with strength \cdraw, who wins when two colluding players
draw, and is inactive whenever a game has at least one non-colluding
player.  Then $H_N\colon\cdraw=\draw$ corresponds to no collusion and
the likelihood function would have

@@

\begin{table}
\centering
\begin{tabular}{l|cccccc}
&\rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Geller} 
&\rotatebox{-90}{Petrosian\ }
&\rotatebox{-90}{Korchnoi} 
&\rotatebox{-90}{Filip}
& $\cdots$
\\ \hline %       Fischer   Geller    Petrosian Korchnoi    Filip
Fischer    (USA)  &   -     & \halfb  & \halfw  & \oneww    &\halfb    &\\
Geller    (USSR)  & \halfw  &    -    & \halfb  & \halfw    & \halfw   &\\
Petrosian (USSR)  & \halfb  & \halfw  &    -    & \halfb    & \halfw   &\\
Korchnoi  (USSR)  & \zerob  & \halfb  & \halfw  &   -       & \oneww   &\\
Filip      (TCH)  & \halfw  & \halfb  & \halfb  & \zerob    &  -       &\\
$\vdots$&&&&&\\
\end{tabular}
\caption{Subset of the Fifth Interzonal round-robin, Stockholm 1962.\label{interzonal}
  Results $\in\{0,\halfw,1\}$ and box colour refer to row
  player.  Thus, for example, Fischer played black against Geller and drew}
\end{table}

The fifth Interzonal took place in Stockholm in 1962, in the form of a
23-player round-robin tournament.  For the results shown in
Table~\ref{interzonal}, we would have a likelihood function

@@

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing white); but the full likelihood
function for the interzonal has 297 terms, including $\draw^{110}$.  To
investigate the possibility of increased Soviet drawing proclivity, we
introduce a Soviet draw monster with strength~$\draw_s$, who wins when two
Soviet players draw, and is inactive if a game has at least one
non-Soviet player.  Then $H_N\colon \draw_s=\draw$ corresponds to no collusion
and the likelihood function for Table~\ref{interzonal} would be

@@

The maximum likelihood estimate is $\draw\simeq 0.043, \draw_s=0.354$; a
constrained maximization gives a likelihood ratio of about 5096, or a
$p$-value of about $3.6\times 10^{-5}$.  Indeed we could go further
and plot a profile likelihood curve for the log-odds ratio
$\log\left(\draw_s/\draw\right)$ which, if Dirichlet, would be accurately
parabolic over a wide range~\cite[page 343]{ohagan2004}.  This is
illustrated in figure~\ref{proflike}, that shows a credible interval
of about 1.8-2.4.

>>>>

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be two-way
competitive situations such as tennis in which one player is
identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
considers~$n$ players and assigns a non-negative number $p_i$ to player
$i$, with $\sum p_i=1$.  Then player $i$ beats player $j$ with
probability $\frac{p_i}{p_i+p_j}$.  This simple system has been
modified and extended many times, for example to account for multi-way
competitions such as {\em MasterChef Australia}~\citep{hankin2017},
and non-competitive situations such as consumer
behaviour~\citep{hankin2010} and mating preference~\citep{west2008}.

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage,
winning about 37.5\% of tournament games (black wins 27.6\%); and
secondly, many chess games, about 39.4\%, are drawn (figures from
Chessbase).

\subsection{Draws} Chess games may be won by either player, or may be
drawn; one inference of a draw is that neither player is better than
the other.  Before 1867, tournament games that were drawn were
replayed until one or other player won---reducing a chess game to a
Bernoulli trial (with probability one).  This system became unworkable
due to the large number of draws, resulting in the modern practice of
awarding each player of a drawn game half a point, keeping chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are usually poorer in spectacular or striking elements such as
sharp tactical combinations or sacrifice.  Excessive drawing dissuades
sponsors, leaves spectators feeling cheated, and gives commentators
less material to discuss.  Players themselves do not seem to value
draws in the wider context of the game: \citet{fischer1969}, for
example, includes only 8 draws in his compilation of 60 memorable
games, surely not representative of tournament play in which about
39.4\% of games are drawn (binomial test, $p<10^{-5}$).

@@

Such arguments suggest that one should infer more from a drawn game
than simple equality of skill: the play itself tends to be different
and more cautious, the drawing players have a different mindset, and
drawn games have less value in the wider context of chess.

\subsection{First-mover advantage} Traditionally White moves first in
chess, and this feature introduces an asymmetry into play.
Historically, we see about 1.35 white wins for every black win.
However, interpretations of this statistic vary and the issue is hotly
debated with opinions ranging from that of~\cite{adams2007}---``White
to play and win'' to~\citet{adorjan1988}---``Black is OK!''.

@@

Here I show how Bradley-Terry may be generalized to account for the
existence of draws and the first-mover advantage.  Consider first the
standard Bradley-Terry model for a finite number of chess players.
Ignoring the effect of playing white, we wish to find numbers
(``strengths") $p_i\geqslant 0$ for $1\leqslant i\leqslant n$ with
$\sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is
$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
assumption of independence of trials, would be

@@

where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess game is regarded as a
three-way competition between the two players and a third entity (a
``draw monster'') that wins if the game is drawn.  The system may be
further modified to account for first-move advantage by positing an
entity whose strength is added to that of the white player.  This
methodology affords insight into players' strengths, response to
playing black or white, and risk-aversion as manifested by probability
of drawing.  If we consider a chess game to be a three-way trial, we
seek a likelihood function for non-negative $p_1,\ldots,p_n,\draw$
(here \draw\ is the draw monster's strength, fraktur font being used
for reified entities' strengths) with $\draw+\sum p_i=1$.  A suitable
likelihood function would be

@@

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength \white\ is
added to the white player.  The strength of the white player is thus
the sum of the player's strength and \white.  Recalling that a game
might be a draw, we obtain the probabilities as given in
Table~\ref{probabilities}.  The corresponding likelihood function is
defined on $\white+\draw+\sum p_i=1$, given below in equation
\ref{likelihood_WD}:

@@

\begin{align}\nonumber
\prod_{i<j}&{}
\left(\frac{p_i+\white}{p_i+p_j+\draw+\white}  \right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+\draw+\white    }\right)^{L_{ij}^\mathrm{white}}\\
&{} \left(\frac{p_i}{p_i+p_j+\draw+\white}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
\left(\frac{p_j+\white}{p_i+p_j+\draw+\white  }\right)^{L_{ij}^\mathrm{black}}\\
&{} \left(\frac{\draw }{p_i+p_j+\draw+\white  }\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
\nonumber
\end{align}

where $W_{ij}^\mathrm{white}$ is the number of games won by $i$ when
playing white against $j$, and so on.  Note that the power of the
final term, $D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}$ is the total
number of games drawn between $i$ and $j$ irrespective of colour.  We
now apply this and related likelihood models to real chess datasets
using the Method of Support~\citep{edwards1972}.  To test any
hypothesis $H$ on likelihood function $\mathcal{L}$ one first
identifies the constraint on hypothesis space corresponding to $H$.
One then finds the unconstrained maximum likelihood
$\mathcal{L}_\mathrm{free}$ and compares it with the maximum
likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization is
constrained by $H$.  The test statistic is then the likelihood ratio
$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
$n$ is the number of degrees of freedom implemented by $H$.

@@

The Method of Support \citep{edwards1972} involves only numerical
optimization (not integration which is computationally expensive).
The analysis presented here uses the framework of the
R~\citep{rcore2019} \pkg{hyper2} package~\citep{hankin2017} as a
convenient and unified interface.  Because gradients are available,
convergence to the evaluate is generally rapid and hypothesis-specific
constraints on parameter space are, in general, linear and
straightforward to implement.  The total runtime for all the tests
presented here is less than a few seconds on a desktop PC.

@@

Grandmasters Anand, Karpov, and Kasparov have long-standing rivalry
and have played many games amongst themselves.  A summary may be found
in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes

@@

\begin{figure}[t]
\begin{center}
<<ddc,echo=FALSE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
  \end{center}
\end{figure}

This likelihood function is easily maximized numerically and the
evaluate is shown visually in Figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $\draw=0$ cannot possibly be the
case: the likelihood for this hypothesis would be identically zero,
owing to the nonzero power of \draw\ in the numerator of
equation~\ref{kka_likelihood}.  From the figure, we see that the
maximum likelihood estimate for draw strength $\hat{\draw}\simeq
0.53$: compare the na\"ive estimate of $\frac{224}{365}\simeq 0.61$
given by observing 224 draws among 265 games.  The ML estimate is
lower because the maximization process automatically adjusts for the
different players' strengths.  Secondly, we may make inferences about
$\white$, the extra strength conferred by playing white: it is
conceivable that $\white=0$ as this term does not appear on its own in
the numerator of any term.  We may test $H_0\colon \white=0$ using the
Method of Support by maximizing the likelihood subject to the
constraint $\white=0$ and comparing with the likelihood when the
optimization is free.

@@

As argued above, players with a cautious style tend to draw and this
is an important and interesting feature of the game (even if drawn
games themselves are boring).  One line of investigation might be a
likelihood function that can account for differences in draw
proclivity among the players.  This may be achieved by considering a
game between two players to be a contest between five entities: the
two real players, the white advantage, and two {\em personalised} draw
entities, one for each real player; a drawn game is counted as a win
for either of the draw monsters.  Under these assumptions,
Table~\ref{kka} translates into the following likelihood function:

@@

Here I consider game results from Cura\c{c}ao 1962, reproduced in
Table~\ref{candidates}.  We see that all twelve of the games between
the three accused players ended in a draw, a significant result given
51 winning games and 54 draws (Fisher's exact test, one-sided,
$p=0.0079$).  To quantify the evidence for any such collusion,
accounting for the player skills, we introduce a new draw entity: a
collusive draw monster with strength \cdraw, who wins when two
colluding players draw, and is inactive whenever a game has at least
one non-colluding player.  Then $H_N\colon\cdraw=\draw$ corresponds to
no collusion and the likelihood function would have

@@

\begin{table}
\centering
\begin{tabular}{l|cccccc}
&\rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Geller} 
&\rotatebox{-90}{Petrosian\ }
&\rotatebox{-90}{Korchnoi} 
&\rotatebox{-90}{Filip}
& $\cdots$
\\ \hline %       Fischer   Geller    Petrosian Korchnoi    Filip
Fischer    (USA)  &   -     & \halfb  & \halfw  & \oneww    &\halfb    &\\
Geller    (USSR)  & \halfw  &    -    & \halfb  & \halfw    & \halfw   &\\
Petrosian (USSR)  & \halfb  & \halfw  &    -    & \halfb    & \halfw   &\\
Korchnoi  (USSR)  & \zerob  & \halfb  & \halfw  &   -       & \oneww   &\\
Filip      (TCH)  & \halfw  & \halfb  & \halfb  & \zerob    &  -       &\\
$\vdots$&&&&&\\
\end{tabular}
\caption{Subset of the Fifth Interzonal round-robin, Stockholm
  1962.\label{interzonal} Results $\in\{0,\halfw,1\}$ and box colour
  refer to row player.  Thus, for example, Fischer played black
  against Geller and drew.  The full dataset, not shown here, has 23
  players}
\end{table}

The fifth Interzonal took place in Stockholm in 1962, in the form of a
23-player round-robin tournament.  For the 20 game results shown in
Table~\ref{interzonal}, we would have a likelihood function

@@

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing white); but the full likelihood
function for the interzonal has 297 terms, including $\draw^{110}$.
To investigate the possibility of increased Soviet drawing proclivity,
we introduce a Soviet draw monster with strength~$\draw_s$, who wins
when two Soviet players draw, and is inactive if a game has at least
one non-Soviet player.  Then $H_N\colon \draw_s=\draw$ corresponds to
no collusion and the likelihood function for the data of
Table~\ref{interzonal} would be

@@

The maximum likelihood estimate for the full 23-player dataset is
$\draw\simeq 0.043, \draw_s=0.354$; a constrained maximization gives a
likelihood ratio of about 5096, or a $p$-value of about $3.6\times
10^{-5}$: there is strong evidence that Soviet players do indeed draw
more frequently than non-Soviets.  Indeed we could go further and plot
a profile likelihood curve for the log-odds ratio
$\log\left(\draw_s/\draw\right)$ which, if Dirichlet, would be
accurately parabolic over a wide range~\cite[page 343]{ohagan2004}.
This is illustrated in figure~\ref{proflike}, that shows a credible
interval of about 1.8-2.4: we can be reasonably sure that
$\draw_s/\draw\geqslant e^{1.8}\simeq 6.05$.
